% !TEX root = ../main.tex

% Exercises section

\section{Exercises}

\begin{enumerate}

\item
This exercise is a worked extended example from Nishihara et al. \cite{Nishihara:2013}. The factor graph from Figure \ref{fig:factorgraph} is shown again here.

\begin{figure}[h]
\begin{center}
\tikz{
\node[latent, xshift=-1cm] (t1) {$\theta_1$};
\node[latent, xshift=1cm] (t2) {$\theta_2$};
\node[latent, yshift=-2cm] (t3) {$\theta_3$};
\node[latent, xshift=2cm, yshift=-2cm] (t4) {$\theta_4$};
\node[obs, xshift=1cm, yshift=-4cm] (t5) {$\theta_5$};
\factor[above=of t1] {f1} {above:Gaussian} {} {t1};
\factor[above=of t2] {f2} {above:Gaussian} {} {t2};
\factor[below=of t1, xshift=1cm] {fplus} {left:sum} {t1,t2} {t3};
\factor[above=of t4] {f4} {above:Gaussian} {} {t4};
\factor[below=of t3, xshift=1cm] {fprod} {left:product} {t3,t4} {t5}
}
\end{center}
\end{figure}

Some factors and the scaling constraints that they impose is provided in the following table.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|}
\hline
factor & constraints \\
\hline
$c=a+b$ & $d_a = d_b = d_c$ \\
$c = a\times b$ & $d_c= d_a + d_b$ \\
$c$ observed & $d_c=0$ \\
\hline
\end{tabular}
\end{table}

\begin{enumerate}

\item
Find the scaling constraint matrix $\mathcal{C}$ for this factor graph.

\begin{solution}
The sum factor adds the rows $d_1 - d_3 = 0$ and $d_2 - d_3 = 0$. \\
The product factor adds the row $d_3 + d_4 - d_5 = 0$. \\
The observed factor adds the row $d_5 = 0$.

Aggregating the rows produces the following constraint matrix.
\[
\mathcal{C} =
\begin{bmatrix}
1 & 0 & -1 & 0 & 0 \\
0 & 1 & -1 & 0 & 0 \\
0 & 0 & 1 & 1 & -1 \\
0 & 0 & 0 & 0 & 1
\end{bmatrix}
\]
\end{solution}

\item
Find the null space of $\mathcal{C}$.

\begin{solution}
A basis for the null space of $\mathcal{C}$ is given by the solution to the following system of equations.
\[
\begin{bmatrix}
1 & 0 & -1 & 0 & 0 \\
0 & 1 & -1 & 0 & 0 \\
0 & 0 & 1 & 1 & -1 \\
0 & 0 & 0 & 0 & 1
\end{bmatrix}
\begin{bmatrix}
d_1 \\ d_2 \\ d_3 \\ d_4 \\ d_5
\end{bmatrix}
=
\begin{bmatrix}
0 \\ 0 \\ 0 \\ 0 \\ 0
\end{bmatrix}
\]
Solving for the solution leads to the null space
\[
\mathcal{N}(\mathcal{C}) = \left\{c
\begin{bmatrix}
1 \\ 1 \\ 1 \\ -1 \\ 0
\end{bmatrix}
: c\in\mathbb{R}
\right\}
\]
\end{solution}

\item
Under the equivalence class formulation, give the definition of the scaling symmetry detector for this factor graph $\mathcal{F}$.

\begin{solution}
The scaling symmetry detector is defined as
\[
\Delta_\mathcal{F}(\theta) = \{\Sigma_d\theta:d\in\mathcal{N}(\mathcal{C})\}
\]
where $\Sigma_d$ is the diagonal matrix with $(e^{d_1},...,e^{d_5})$ on the diagonal and $\mathcal{N}(\mathcal{C})$ is as found in the previous question. 
\end{solution}

\item
Given $\theta^*=(\theta_1^*,...,\theta_5^*)^T$, find $[\theta^*]$.

\begin{solution}
The equivalence class of $\theta^*$ can be found using the symmetry detector defined in the previous question.
\[
[\theta^*] = \Delta_\mathcal{F}(\theta^*) = \{\Sigma_d\theta^*:d\in\mathcal{N}(\mathcal{C})\} = \left\{
\begin{bmatrix}
e^c\theta_1^* \\
e^c\theta_2^* \\
e^c\theta_3^* \\
e^{-c}\theta_4^* \\
\theta_5^*
\end{bmatrix}
:
c\in\mathbb{R}
\right\}
\]
\end{solution}

\iffalse
\item
Let $\|\cdot\|$ denote the Euclidean norm. Suppose that for any equivalence class $[\theta]$, no two members have the same norm. Consider a symmetry breaker of the form
\[
\phi_\mathcal{F}([\theta])=\{\theta:\theta\in[\theta],\|\theta\|=1\}
\]
\fi

\end{enumerate}

\iffalse
\item
Consider the equivalence class formulation. Suppose that a factor graph $\mathcal{F}$ has a scaling symmetry such that every equivalence class $[\theta]$ has a member $\theta^*$ with Euclidean norm $\|\theta^*\|=1$. Define
\[
\phi_\mathcal{F}([\theta])=\{\theta:\theta\in[\theta],\|\theta\|=1\}
\]
Show that $\phi_\mathcal{F}([\theta])$ is a singleton for any $[\theta]$, i.e., that $\phi_\mathcal{F}$ is a scaling symmetry breaker. (\textit{Remark: here we use a singleton to denote the representative to allow this $\phi_\mathcal{F}$ to be defined in simple notation.})

\begin{solution}
We show this by contradiction.

Suppose that for some equivalence class $[\theta]$, $\phi_\mathcal{F}([\theta])$ returns a set containing two members $\theta^*$ and $\theta$. By definition of $[\theta]$, we must have $\theta^*=\Sigma_{d^*}\theta$ for some $d^*\in\mathcal{N}(\mathcal{C})$. Then
\[
\|\theta^*\| = \|\Sigma_{d^*}\theta\| = 1 = \|\theta\| \iff \Sigma_{d^*}^T\Sigma_{d^*} = \mathcal{I}
\]
where $\mathcal{I}$ is the identity matrix. Hence $\Sigma_{d^*}$ must be orthogonal. But because $\Sigma_{d^*}$ is diagonal and positive definite, $\Sigma_{d^*}$ must be the identity matrix and so $\theta^*=\mathcal{I}\theta = \theta$. Contradiction and thus $\phi_\mathcal{F}$ returns a singleton for any $[\theta]$.

We note that this $\phi_\mathcal{F}$ is only well-defined under the assumption that every equivalence class has some $\theta$ with unit norm. This is generally not true in all factor graphs. For example, a factor graph that has no scaling symmetries has every $\theta$ is in its own equivalence class but $\theta$ may not have a unit norm.
\end{solution}
\fi

\newpage

\item
Consider the following factor graph that has an exponential prior with rate $\beta = \frac{1}{2}$ and a chi-squared prior with degrees of freedom $d=2$:

\begin{figure}[h]
\begin{center}
\tikz{
\node[latent, xshift=-1cm] (t1) {$\theta_1$};
\node[latent, xshift=1cm] (t2) {$\theta_2$};
\node[latent, yshift=-2cm] (other) {...};
\factor[above=of t1] {f1} {above:Exp$\left(\frac{1}{2}\right)$} {} {t1};
\factor[above=of t2] {f2} {above:$\chi^2(2)$} {} {t2};
\factor[below=of t1,xshift=1cm] {fplus} {left:sum} {t1,t2} {other}
}
\end{center}
\end{figure}

This factor graph has a permutation symmetry. Explain why and suggest one way to break the symmetry under the reparameterization formulation.

Let $\lambda$ denote the Lebesgue measure. You may find the following useful:
\[
\mathbb{P}(\theta_1\in dx) = \lambda(dx)\beta e^{-\beta x} \qquad x\in\mathbb{R}_+
\]
\[
\mathbb{P}(\theta_2\in dx) = \lambda(dx)\frac{1}{2^{\frac{d}{2}}\Gamma\left(\frac{d}{2}\right)}x^{\frac{d}{2}-1}e^{-\frac{x}{2}} \qquad x\in\mathbb{R}_+
\]

\begin{solution}
Notice that
\[
\mathbb{P}(\theta_1\in dx) = \lambda(dx)\frac{1}{2}e^{-\frac{1}{2}x} \qquad x\in\mathbb{R}_+
\]
and
\[
\mathbb{P}(\theta_2\in dx) = \lambda(dx)\frac{1}{2^{\frac{2}{2}}\Gamma\left(\frac{2}{2}\right)}x^{\frac{2}{2}-1}e^{-\frac{x}{2}} = \lambda(dx)\frac{1}{2}e^{-\frac{x}{2}} \qquad x\in\mathbb{R}_+
\]
and so $\theta_1$ and $\theta_2$ have the same distribution. Hence the permutation symmetry exists because the inputs to the sum factor are symmetric. In this case, the symmetry can be broken by merging the sum factor with its inputs as the sum is specified by a gamma distribution with shape $\alpha=2$ and rate $\beta=\frac{1}{2}$. We show this using the Laplace transform. For $r\in\mathbb{R}_+$, we have
\[
\mathbb{E}[e^{-r\theta_1}] = \mathbb{E}[e^{-r\theta_2}] = \frac{2^{-1}}{2^{-1}+r}
\]
Then by independence of $\theta_1$ and $\theta_2$,
\[
\mathbb{E}[e^{-r(\theta_1+\theta_2)}] = \mathbb{E}[e^{-r\theta_1}]\mathbb{E}[e^{-r\theta_2}] = \left(\frac{2^{-1}}{2^{-1}+r}\right)^2
\]
which corresponds to the Laplace transform of the mentioned gamma distribution. The new factor graph obtained from merging is

\begin{figure}[h]
\begin{center}
\tikz{
\node[latent] (t12) {$\theta_{12}$};
\node[latent, yshift=-1.25cm] (other) {...};
\factor[above=of t12] {f1} {above:Gamma$\left(2,2^{-1}\right)$} {} {t12};
\edge {t12} {other}
}
\end{center}
\end{figure}
\end{solution}

\end{enumerate}